---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
    toc: true
    number_sections: true
---


```{r setup_rr-pa1, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, collapse = TRUE,
                      fig.path = "figure/")
```


# Loading and preprocessing the data

Let's use the `read_csv()` function from the `readr` package to read the data in.

```{r loading, message=FALSE}
library(readr)

activity <- read_csv("activity.zip")

activity

sapply(activity, class)
```

As we can see, the `read_csv()` function:

  - Handles compressed files.  
  - Returns a `tibble` with:  
      - the `date` variable already formatted as variable of class `Date`;  
      - the `steps` and `interval` variables as `numeric` variables.  


# What is mean total number of steps taken per day?

As requested by the assignment instructions, in this section we'll ignore *missing values* and proceed as follows.

## Make a histogram of the total number of steps taken each day

First, we **summarize the total number of steps taken per day**.

```{r stepsPerDay, message=FALSE}
library(dplyr)

stepsPerDay <- activity %>% 
        na.omit %>%                       ## Ignoring NAs.
        group_by(date) %>% 
        summarize(total = sum(steps))

stepsPerDay
```

Then, we make a **histogram of the total number of steps per day**.

```{r stepsPerDay_hist}
hist(stepsPerDay$total, col = "steelblue",
     xlab = "Steps per day", ylab = "Count",
     main = "Total number of steps per day histogram")
```

## Calculate and report the mean and median total number of steps taken per day

We can use the `summary()` function to calculate the ***mean*** and ***median*** **total number of steps taken per day**.

```{r stepsPerDay_smries}
stepsPerDay_smries <- summary(stepsPerDay$total)
stepsPerDay_smries
```

As we can see, the *mean* and the *median* are practically the same:

  - The **mean total number of daily steps** is **`r as.integer(stepsPerDay_smries["Mean"])`**.  
  - The **median total number of daily steps** is **`r as.integer(stepsPerDay_smries["Median"])`**.  


# What is the average daily activity pattern?

## Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

In order to make the requested time series plot, we start by **summarizing the average steps per intervals**.

```{r stepsPerIntervals, message=FALSE}
library(dplyr)

stepsPerIntervals <- activity %>% 
        na.omit %>%                       ## Ignoring NAs.
        group_by(interval) %>% 
        summarize(mean = mean(steps))

stepsPerIntervals
```

So here's a **time series plot** of the average steps taken in each of the five minutes intervals, in which each day of the two month period is divided in.

```{r stepsPerIntervals_timeSeriesPlot}
with(stepsPerIntervals, plot(interval, mean, type = "l",
                             xlab = "Five minutes intervals",
                             ylab = "Average steps",
                             main = "Average steps per daily 5-minute interval"))
```

## Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

To answer this question:

 - We'll first find the index of the maximum avarage number of steps taken in a 5-minute interval.  
 - And then use that index to subset the corrisponding interval in which that maximum occurs.  

```{r maxStepsInterval}
stepsPerIntervals_maxIdx <- which.max(stepsPerIntervals$mean)
maxStepsInterval <- stepsPerIntervals$interval[stepsPerIntervals_maxIdx]
maxStepsInterval
```

So, the daily 5-minute interval where the average maximum number of steps takes place is **the interval `r maxStepsInterval`**.


# Imputing missing values

## Calculate and report the total number of missing values in the dataset

To accomplish this task, we can apply the `summary()` function to our entire datset.

```{r activity_NAs}
activity_smries <- summary(activity)
activity_smries
```

This shows us that the **only variable that contains missing values** is the `steps` variable, with a **total number of `r sum(is.na(activity$steps))` NA's**.

## Devise a strategy for filling in all of the missing values in the dataset

Let's fill in the missing steps values in a given 5-minutes interval, with the **mean for that 5-minutes interval**.

  - So, this strategy assumes that the subject mantains the same behaviour on average.  
  
  - Consequently, we expect that the new average values that we'll explore basically will not differ from those previously calculated.  

**To do so**, we're going to:

  - Group our orginal dataset by the variable `interval`.  
  
  - Create a new column, called `interval.mean`, containing the mean of each of the 5-minute intervals in which all days of the two month period are divided in.  
  
  - Finally, fill in the missing values of the `steps` variable with the corrisponding `interval.mean` values.  

```{r activity_filledIn, message=FALSE}
library(dplyr)

activity_filledIn <- activity %>% 
        group_by(interval) %>% 
        mutate(interval.mean = mean(steps, na.rm = TRUE)) %>% 
        mutate(steps = ifelse(is.na(steps), interval.mean, steps))

activity_filledIn
```

## Create a new dataset that is equal to the original dataset but with the missing data filled in

At this point we just have to replace the `steps` column of the original `activity` dataset, with the `steps` column from the `activity_filledIn` dataset.

```{r activity_new, message=FALSE}
library(dplyr)

activity_new <- activity %>% 
        mutate(steps = activity_filledIn$steps)

activity_new

sum(is.na(activity_new$steps))            ## No more NAs.
```

## Make a histogram of the total number of steps taken each day, and calculate and report the mean and median total number of steps taken per day

We'll reproduce with our new dataset (`activity_new`), the same steps we did before with the original dataset (`activity`):

  - First, **summarize the total number of steps taken per day**.

```{r stepsPerDay_new, message=FALSE}
library(dplyr)

stepsPerDay_new <- activity_new %>% 
        group_by(date) %>% 
        summarize(total = sum(steps))

stepsPerDay_new
```

  - Then, make the **histogram of the total number of steps per day**.

```{r stepsPerDay_new_hist}
hist(stepsPerDay_new$total, col = "steelblue",
     xlab = "Steps per day", ylab = "Count",
     main = "Total number of steps per day histogram")
```

  - Calculate and report the **mean and median total number of steps taken per day**.

```{r stepsPerDay_new_smries}
stepsPerDay_new_smries <- summary(stepsPerDay_new$total)
stepsPerDay_new_smries
```

### Do these values differ from the estimates from the first part of the assignment?

As we expected, the **mean and median total number of daily steps** obtained in the new dataset:

  - **mean**: **`r as.integer(stepsPerDay_new_smries["Mean"])`**;  
  - **median**: **`r as.integer(stepsPerDay_new_smries["Median"])`**;  

***are practically the same*** from those we've calculated before from the original dataset.

**However**, as the **proportion of missing values** within the hole dataset **(`r round(mean(is.na(activity$steps))*100)` percent)** is ***not entirely irrelevant***, another imputing missing values strategy would've altered these average values with some relevant degree of incidence.

### What is the impact of imputing missing data on the estimates of the total daily number of steps?

Both histograms show us that:

  - the data distribution remains unchanged;  
  
  - and that the impact of our imputing missing data strategy is expressed in a uniformly proportional growth of the total daily number of steps.  


# Are there differences in activity patterns between weekdays and weekends?

## Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day

Let's create this new factor variable in the new filled-in missing values dataset (`activity_new`), using the `mutate()` function and calling it `type.of.day`.

```{r patterns}
library(dplyr)

activity_new <- activity_new %>% 
        mutate(type.of.day = factor(ifelse(weekdays(date)=="sábado" |
                                           weekdays(date)=="domingo",
                                           "weekend", "weekday"),
                                    levels = c("weekday", "weekend")))
activity_new

table(activity_new$type.of.day)
```

## Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)

 - Let's begin **grouping by `type.of.day` and `interval`, and summarize the average steps**.  

```{r stepsPerTypeOfDays, message=FALSE}
library(dplyr)

stepsPerTypeOfDays <- activity_new %>% 
        group_by(type.of.day, interval) %>% 
        summarize(mean = mean(steps))

stepsPerTypeOfDays
```

  - Here's the requested **panel plot**.  

```{r stepsPerIntervals_panelPlot}
library(lattice)

xyplot(mean ~ interval | type.of.day, data = stepsPerTypeOfDays,
       layout = c(1, 2),
       type = "l",
       xlab = "Five minutes intervals",
       ylab = "Average steps",
       main = "Average steps per daily intervals on weekdays and weekends")
```

As we can see:

  - There are more average steps in the first intervals of the day on weekdays, than on weekends (the subject begins to move earlier on weekdays).  
  
  - The average steps in the following intervals of the day are bigger on weekends, than they are on weekdays (the subject tends to move more during the rest of the day on weekends).  
  
***
